---
title: "Introduction to R and RStudio"
author: "Julie Lowndes"
date: "April 15, 2016"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

Here is an analogy to start us off. **If you were a pilot, R is an an airplane.** You can use R to go places! With practice you'll gain skills and confidence; you can fly further distances and get through tricky situations. You will become an awesome pilot and can fly your plane anywhere. 

And **if R were an airplane, RStudio is the airport**. RStudio provides support! Runways, communication and other services, and just makes your overall life easier. So although you can fly your plane without an airport and we could learn R without RStudio, that's not what we're going to do. 

> We are learning R together with RStudio and its many supporting features. 

Something important to mention is that you are learning a new language here. It's an ongoing process, it takes time, you'll make mistakes, it can be frustrating, but it will be overwhelmingly awesome in the long run. We all speak at least one language; no matter how fluent you are, you'll always be learning, you'll be trying things in new contexts, etc. There are miscommunications and moments of clarity, but you are overall communicating, and you're better off because of it. 


https://www.mozillascience.org/introducing-mozilla-science-study-groups


To learn R and RStudio we will be using Dr. Jenny Bryan's lectures from STAT545 at UBC. I have modifed them slightly here for our purposes; to see them in their full and awesome entirety, visit [stat545-ubc.github.io](https://stat545-ubc.github.io/). Specifically, we'll be using these lectures: 

- [R basics, workspace and working directory, RStudio projects](http://stat545-ubc.github.io/block002_hello-r-workspace-wd-project.html)
- [Basic care and feeding of data in R](http://stat545-ubc.github.io/block006_care-feeding-data.html)

Something we won't cover today but that will be helpful to you in the future is:

- [The many flavors of R objects](http://stat545-ubc.github.io/block004_basic-r-objects.html)

I've modified them in part with my own text and in part with text from Software Carpentry's [R for reproducible scientific analysis](http://swcarpentry.github.io/r-novice-gapminder/), specifically: 

- [Seeking help](http://swcarpentry.github.io/r-novice-gapminder/03-seeking-help.html)


# R basics, workspace and working directory, RStudio projects
*(modified from Jenny Bryan's [STAT545](http://stat545-ubc.github.io/block002_hello-r-workspace-wd-project.html))*

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## R at the command line, RStudio goodies

Launch RStudio/R.

Notice the default panes:

  * Console (entire left)
  * Environment/History (tabbed in upper right)
  * Files/Plots/Packages/Help (tabbed in lower right)

FYI: you can change the default location of the panes, among many other things: [Customizing RStudio](https://support.rstudio.com/hc/en-us/articles/200549016-Customizing-RStudio).

Go into the Console, where we interact with the live R process.

Make an assignment and then inspect the object you just created.

```{r}
x <- 3 * 4
x
```

All R statements where you create objects -- "assignments" -- have this form:
```{r eval = FALSE}
objectName <- value
```
and in my head I hear, e.g., "x gets 12".

You will make lots of assignments and the operator `<-` is a pain to type. Don't be lazy and use `=`, although it would work, because it will just sow confusion later. Instead, utilize **RStudio's keyboard shortcut: Alt + - (the minus sign)**.

Notice that RStudio automagically surrounds `<-` with spaces, which demonstrates a useful code formatting practice. Code is miserable to read on a good day. Give your eyes a break and use spaces.

RStudio offers many handy [keyboard shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts). Also, Alt+Shift+K brings up a keyboard shortcut reference card.

Object names cannot start with a digit and cannot contain certain other characters such as a comma or a space. You will be wise to adopt a [convention for demarcating words](http://en.wikipedia.org/wiki/Snake_case) in names.

```
i_use_snake_case
other.people.use.periods
evenOthersUseCamelCase
```

Make an assignment
```{r}
this_is_a_really_long_name <- 2.5
```

Make another assignment
```{r}
this_is_shorter <- 2 ^ 3
```

To inspect this, try out RStudio's completion facility: type the first few characters, press TAB, add characters until you disambiguate, then press return.
```{r}
this_is_shorter
```

One more:
```{r}
jenny_rocks <- 2
```

Let's try to inspect:
```{r}
jennyrocks
jeny_rocks
```

Implicit contract with the computer / scripting language: Computer will do tedious computation for you. In return, you will be completely precise in your instructions. Typos matter. Case matters. Get better at typing.

Remember that this is a language, not unsimilar to English! There are times you aren't understood -- your friend might say 'what?' but R will say 'error'. 


## R functions, help pages

R has a mind-blowing collection of built-in functions that are accessed like so

```{r eval = FALSE, tidy = FALSE}
functionName(arg1 = val1, arg2 = val2, and so on)
```

Let's try using `seq()` which makes regular sequences of numbers and, while we're at it, demo more helpful features of RStudio.

Type `se` and hit TAB. A pop up shows you possible completions. Specify `seq()` by typing more to disambiguate or using the up/down arrows to select. Notice the floating tool-tip-type help that pops up, reminding you of a function's arguments. If you want even more help, press F1 as directed to get the full documentation in the help tab of the lower right pane. 

Type the arguments `1, 10` and hit return. 
```{r}
seq(1, 10)
```

We could probably infer that the `seq()` function makes a sequence, but let's learn for sure. Type (and you can autocomplete) and let's explore the help page:

```{r}
?seq 
help(seq) # same as ?seq
```

The above also demonstrates something about how R resolves function arguments. You can always specify in `name = value` form. But if you do not, R attempts to resolve by position. So above, it is assumed that we want a sequence `from = 1` that goes `to = 10`. Since we didn't specify step size, the default value of `by` in the function definition is used, which ends up being 1 in this case. For functions I call often, I might use this resolve by position for the first
argument or maybe the first two. After that, I always use `name = value`.


The help page tells the name of the package in the top left, and broken down into sections:

 - Description: An extended description of what the function does.
 - Usage: The arguments of the function and their default values.
 - Arguments: An explanation of the data each argument is expecting.
 - Details: Any important details to be aware of.
 - Value: The data the function returns.
 - See Also: Any related functions you might find useful.
 - Examples: Some examples for how to use the function.

The examples can be copy-pasted into the console for you to understand what's going on. Let's try it. 

> Exercise: Look up the help file for a function you know. For example, `?getwd()`, `?plot()`.


Help for when you only sort of remember the function name: double-questionmark:
```{r}
??install 
```


Not all functions have (or require) arguments:
```{r}
date()
```

Now look at your workspace -- in the upper right pane. The workspace is where user-defined objects accumulate. You can also get a listing of these objects with commands:

```{r}
objects()
ls()
```

If you want to remove the object named `y`, you can do this

```{r}
rm(y)
```

To remove everything:

```{r}
rm(list = ls())
```

or click the broom in RStudio's Environment pane.

> Exercise: Clear your workspace, then create a few new variables. Discuss what makes a good filename.

## History, working directory, scripts 

One day you will need to quit R, go do something else and return to your analysis later.

One day you will have multiple analyses going that use R and you want to keep them separate.

One day you will need to bring data from the outside world into R and send numerical results and figures from R back out into the world.

To handle these real life situations, you need to make two decisions:

  * What about your analysis is "real", i.e. you will save it as your lasting record of what happened?

  * Where does your analysis "live"?
  

As a beginning R user, it's OK to consider your workspace "real". _Very soon_, I urge you to evolve to the next level, where you consider your saved R scripts as "real". (In either case, of course the input data is very much real and requires preservation!) With the input data and the R code you used, you can reproduce
_everything_. You can make your analysis fancier. You can get to the bottom of puzzling results and discover and fix bugs in your code. You can reuse the code to conduct similar analyses in new projects. You can remake a figure with different aspect ratio or save is as TIFF instead of PDF. You are ready to take questions. You are ready for the future.

We're not going to cover workspaces today, but you can learn about it in this RStudio article: [Working Directories and Workspaces](https://support.rstudio.com/hc/en-us/articles/200711843-Working-Directories-and-Workspaces).


### Working directory

Any process running on your computer has a notion of its "working directory". In R, this is where R will look, by default, for files you ask it to load. It also where, by default, any files you write to disk will go. Chances are your current working directory is the directory we inspected above, i.e. the one where RStudio wanted to save the workspace.

You can explicitly check your working directory with:

```{r}
getwd()
```

It is also displayed at the top of the RStudio console.

As a beginning R user, it's OK let your home directory or any other weird directory on your computer be R's working directory. _Very soon_, I urge you to evolve to the next level, where you organize your analytical projects into directories and, when working on project A, set R's working directory to the associated directory.

__Although I do not recommend it__, in case you're curious, you can set R's working directory at the command line like so. You could also do this in a script.

```{r eval = FALSE}
setwd("~/myCoolProject")
```


But there's a better way. A way that also puts you on the path to managing your R work like an expert.

## RStudio projects

Keeping all the files associated with a project organized together -- input data, R scripts, analytical results, figures -- is such a wise and common practice that RStudio has built-in support for this via its _projects_.

[Using Projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)

Let's make one to use for the rest of this workshop/class. Do this: File > New Project .... The directory name you choose here will be the project name. Call it whatever you want (or follow me for convenience).

I created a directory and, therefore RStudio project, called `swc` in my `github` directory, FYI.


```{r eval=FALSE}
setwd("~/github/swc")
```

Now check that the "home" directory for your project is the working directory of our current R process:

```{r}
getwd()
# "/Users/julialowndes/github/swc" 
```
_I can't print my output here because this document itself does not reside in the RStudio Project we just created._

This is the absolute path, just like we learned in the shell this morning. But from here, your paths within this project can be relative, and so our files within our project could work on your computer or mine, without worrying about the absolute paths.

Let's enter a few commands in the Console, as if we are just beginning a project. `cars` is a dataset loaded into R.
```{r eval=FALSE}
cars
```

```{r toy-line}
plot(cars)  
z <- line(cars)
abline(coef(z), col = "purple")
dev.print(pdf, "toy_line_plot.pdf")
```

## Our first script!

Let's say this is a good start of an analysis and your ready to start preserving the logic and code. Visit the History tab of the upper right pane. Select these commands. Click "To Source". Now you have a new pane containing a nascent R script. Click on the floppy disk to save. Give it a name ending in `.R` or `.r`, I used `toy-line.r` and note that, by default, it will go in the directory associated with your project. It is traditional to save R scripts with a `.R` or `.r` suffix. 

A few things: 

  * Let's comment our script: Comments start with one or more `#` symbols. Use them. RStudio helps you (de)comment selected lines with Ctrl+Shift+C (windows and linux) or Command+Shift+C (mac).

  * Walk through line by line by keyboard shortcut (command + enter) or mouse (click Run in the upper right corner of editor pane).
  
  * Source the entire document -- equivalent to entering `source('toy-line.r')` in the Console -- by keyboard shortcut (shift command S) or mouse (click Source in the upper right corner of editor pane or select from the mini-menu accessible from the associated down triangle).
  
```{r toy-line-commented, eval=FALSE}

## this script plots R's cars data with a fitted line ----

plot(cars)  
z <- line(cars)
abline(coef(z), col = "purple")

## save as .pdf
dev.print(pdf, "toy_line_plot.pdf")
```
Notice that the notation with `----` in a comment also enables us to 'jump' to it in RStudio

This workflow will serve you well in the future:

  * Create an RStudio project for an analytical project
  * Keep inputs there (we'll soon talk about importing)
  * Keep scripts there; edit them, run them in bits or as a whole from there
  * Keep outputs there (like the PDF written above)

Avoid using the mouse for pieces of your analytical workflow, such as loading a dataset or saving a figure. Terribly important for reproducility and for making it possible to retrospectively determine how a numerical table or PDF was actually produced (searching on local disk on filename, among `.R` files, will lead to the relevant script).

```{r include = FALSE}
file.remove(c("toy_line_plot.pdf"))
```



# Basic care and feeding of data in R
*(modified from Jenny Bryan's [STAT545](http://stat545-ubc.github.io/block006_care-feeding-data.html))*


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Let's start fresh. 

You should clean out your workspace. In RStudio, click on the "Clear" broom icon from the Environment tab or use *Session > Clear Workspace*. You can also enter `rm(list = ls())` in the Console to accomplish same.

Now restart R. This will ensure you don't have any packages loaded from previous calls to `library()`. In RStudio, use *Session > Restart R*. Otherwise, quit R with `q()` and re-launch it.

Why do we do this? So that the code you write is complete and re-runnable. If you return to a clean slate often, you will root out hidden dependencies where one snippet of code only works because it relies on objects created by code saved elsewhere or, much worse, never saved at all. Similarly, an aggressive clean slate approach will expose any usage of packages that have not been explicitly loaded. 

Let's check our working directory: `getwd()`

Finally, let's create a new R script from scratch. We will evelop and run our code from there. We'll be using this script today and tomorrow. 

In RStudio, use *File > New File > R Script*. Save this script with a name ending in `.r` or `.R`, containing no spaces or other funny stuff, and that evokes whatever it is we're doing today. Example: `r-software-carpentry-ucsb.r`. 

## Meet your first data.frame: gapminder

We will work with some of the data from the [Gapminder project](http://www.gapminder.org). Jenny Bryan has also released this as an R package, so you could also install it from CRAN and load it into R like so:`install.packages("gapminder"); library(gapminder)`. But here we will use `read.csv` with the file we downloaded before class. 
```{r}
## read gapminder csv
gapminder = read.csv('data/gapminder-FiveYearData.csv')
```


Let's explore: 
```{r eval=FALSE}
## explore the gapminder dataset
gapminder # this is a long data.frame! Let's inspect in different ways
```

```{r}
str(gapminder) #displays the structure of an object
```
`str()` will provide a sensible description of almost anything: when in doubt, just `str()` some of the recently created objects to get some ideas about what to do next.

Whenever you have rectangular, spreadsheet-y data, your default data receptacle in R is a data.frame. Do not depart from this without good reason. 

  * data.frames package related variables neatly together,
    - applying any filtering of observations uniformly
  * most functions, including the latest and greatest packages actually __require__ that your data be in a data.frame
  * data.frames -- unlike general arrays or, specifically, matrices in R -- can hold variables of different flavors such as character data (names), quantitative data (species count), and categorical information (male vs. female)

Let's explore more. 
```{r}
head(gapminder) # shows first 6
tail(gapminder) # shows last 6

head(gapminder, 10) # shows first X that you indicate
tail(gapminder, 12) # guess what this does!

```

More ways to query basic info on a data.frame. 
```{r}
names(gapminder)
dim(gapminder)    # ?dim dimension
ncol(gapminder)   # ?ncol number of columns
nrow(gapminder)   # ?nrow number of rows
length(gapminder) # ?length length; although better for vectors
```

A statistical overview can be obtained with `summary()`

```{r}
summary(gapminder)
```

### Discussion
What other information would you want to know when first exploring new data? 
...How about plotting to see if these data make sense/have outliers?

Although we haven't begun our formal coverage of visualization yet, it's so important for smell-testing dataset that we will make a few figures anyway. Here we use only base R graphics, which are very basic.

```{r first-plots-base-R}
## plot gapminder
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
```

Let's go back to the result of `str()` to talk about data.frames and vectors in R
```{r}
str(gapminder)
```

A data.frame is a special case of a *list*, which is used in R to hold just about anything. data.frames are the special case where the length of each list component is the same. data.frames are superior to matrices in R because they can hold vectors of different flavors (heuristic term explained below), e.g. numeric, character, and categorical data can be stored together. This comes up alot.

### Look at the variables inside a data.frame

To specify a single variable from a data.frame, use the dollar sign `$`. 

Let's explore the numeric variable for life expectancy.
```{r histogram-lifeExp}
## explore numeric variable
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

The year variable is a numeric integer variable, but since there are so few unique values it also functions a bit like a categorical variable.
```{r}
## explore numeric variable that functions like a categorical variable
head(gapminder$year)
summary(gapminder$year)
hist(gapminder$year) # not so interesting
table(gapminder$year) # look at it another way, still not so interesting
```

The variables for country and continent hold truly categorical information, which is stored as a *factor* in R.
```{r}
## explore factor variable
str(gapminder) # ah yes, $continent is a factor.
class(gapminder$continent) # you could also tell with class()
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
hist(gapminder$continent) # whaaaa!?
```

The __levels__ of the factor `continent` are "Africa", "Americas", etc. and this is what's usually presented to your eyeballs by R. In general, the levels are friendly human-readable character strings, like "male/female" and "control/treated". But *never ever ever* forget that, under the hood, R is really storing integer codes 1, 2, 3, etc. Look at the result from `str(gapminder$continent)` if you are skeptical.
```{r}
str(gapminder$continent)
```

This [Janus](http://en.wikipedia.org/wiki/Janus)-like nature of factors means they are rich with booby traps for the unsuspecting but they are a necessary evil. I recommend you resolve to learn how to [properly care and feed for factors](http://stat545.com/block014_factors.html). The pros far outweigh the cons. Specifically in modelling and figure-making, factors are anticipated and accommodated by the functions and packages you will want to exploit.

Here we count how many observations are associated with each continent and, as usual, try to portray that info visually. This makes it much easier to quickly see that African countries are well represented in this dataset.
```{r tabulate-continent}
table(gapminder$continent)
barplot(table(gapminder$continent))
```

> Exercise: Explore `gapminder$gdpPercap`. What commands do you use? 


In the figures below, we see how factors can be put to work in figures. The `continent` factor is easily mapped into "facets" or colors and a legend by the `ggplot2` package. *Making figures with `ggplot2` is covered tomorrow so feel free to just sit back and enjoy these plots or blindly copy/paste.*
```{r factors-nice-for-plots, fig.show = 'hold', out.width = '49%'}
## illustrative plot with ggplot2 (we'll learn this tomorrow)
library(ggplot2) # install.packages("ggplot2")
p <- ggplot(subset(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # initializes stuff
p <- p + scale_x_log10() # log the x axis the right way
p + geom_point() # scatterplot
p + geom_point(aes(color = continent)) # map continent to color
p + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
p + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
  geom_smooth(lwd = 1.5, se = FALSE)
```


## `subset()` to isolate bits of data.frames

There is a stand-alone function called `subset()`, that can isolate pieces of an object for inspection or assignment. `subset()`'s main argument is also (unfortunately) called `subset`.
```{r}
## subset gapminder
subset(gapminder, subset = country == "Uruguay") # Ah, inspecting Uruguay. Self documenting!
```

Contrast the above command with this one accomplishing the same thing:

```{r}
gapminder[1621:1632, ] # No idea what we are inspecting. Don't do this.
```

Yes, these both return the same result. But the second command is horrible for these reasons:

  * It contains [Magic Numbers](http://en.wikipedia.org/wiki/Magic_number_(programming)). The reason for keeping rows 1621 to 1632 will be non-obvious to someone else and that includes __you__ in a couple of weeks.
  * It is fragile. If the rows of `gapminder` are reordered or if some observations are eliminated, these rows may no longer correspond to the Uruguay data.

In contrast, the first command, using `subset()`, is **self-documenting**; one does not need to be an R expert to take a pretty good guess at what's happening. It's also more robust. It will still produce the correct result even if `gapminder` has undergone some reasonable set of transformations (what if it were in in reverse alphabetical order?)

You can use `subset =` and `select =` together to simultaneously filter rows and columns or variables.
```{r}
subset(gapminder, subset = country == "Mexico", 
       select = c(country, year, lifeExp)) # ?c: combines values into a vector or list
```

> Exercise:  
1. Get data just from 2007.   
2. Assign this to a variable.   
3. How many rows are there?   
4. How do you know?  
```{r exercise-answer, eval=FALSE}
# one potential exercise answer, no peeking
x = subset(gapminder, subset = year == 2007) 
str(x)
```

## nested functions and magrittr %>%

Let's return to our barplot call and talk about nested functions. This command has 2 functions: they happen from the inside-out like mathmatical order of operations (PEMDAS). We can take advantage of white space in scripts to help look at these. RStudio has automatic indentation to help. 
```{r eval=FALSE}
barplot(table(gapminder$continent)) # nested, hard to read

barplot(
  table(gapminder$continent)) # whitespace/indentation - slightly easier to read
```

But you can do a lot more than 2: 
```{r eval=FALSE}
barplot(table(sample(head(gapminder$continent, 100)))) # getting pretty miserable

barplot(
  table(
    sample(
      head(gapminder$continent, 100)))) # a lot of ))) to keep track of
```


This can get quite confusing when there multiple functions with multiple arguments. I won't even try to write it without whitespace.  
```{r eval=FALSE}
barplot(
  table(
    sample(
      head(gapminder$continent, 100), # the data, hooray! 
      replace = FALSE, 
      prob    = NULL)),               # a lot of ))'s and ,'s
  xlab = 'continents',                # yikes, these are barplot arguments!
  ylab = 'count') 
```

A few confusing things about nesting at a glance: 

- the actual data we're working with (gapminder$continent) is not immediately obvious
- a lot of )) closing parentheses to keep track of
- a lot of possibility for error


Enter `magrittr`'s "then" operator: %>%

`%>%` which transfers the output on the left to the first argument of the function on the right. This is very powerful when chaining together many operations on the same data frame. We're able to efficiently write out operations, get past the previous problem of multiple variable names without the obfuscation of nesting. 
```{r eval=FALSE}
library(magrittr) # install.packages('magrittr')

gapminder$continent %>%    # start with the data
  head(100) %>%            # before this was head(data, 100); %>% passes data 
  sample(replace = FALSE,  # arguments nicely with the function name
         prob    = NULL) %>%
  table() %>%
  barplot(xlab = 'continents', 
          ylab = 'count')  # arguments nicely with the function name

```

Benefits to %>%

- readable like a story
- indentation makes it easier on the eyes
- fewer commas, multi-parentheses 
- arguments are associated with their function

> Exercise: Use `%>%` and `head()` to visualize `gapminder$gdpPercap`. How would you plot it?


## clean up and save your .r script

OK, let's clean up and save your `.r` script, we'll be using it again tomorrow! Restart R. This will ensure you don't have any packages loaded from previous calls to `library()`. In RStudio, use *Session > Restart R*. Otherwise, quit R with `q()` and re-launch it.

Run through each line of code again, make sure your comments are good, delete anything you don't need. Your script might look like this: 

```{r eval=FALSE}
## explore the gapminder dataset ----
gapminder = read.csv('data/gapminder-FiveYearData.csv')
str(gapminder) #displays the structure of an object
head(gapminder) # shows first 6 by default
tail(gapminder, 12)# shows last X that you indicate, or 6 by default
names(gapminder)
dim(gapminder)    # ?dim dimension
ncol(gapminder)   # ?ncol number of columns
nrow(gapminder)   # ?nrow number of rows
length(gapminder) # ?length length; although better for vectors
summary(gapminder)

## plot gapminder
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)

## explore numeric variable
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)

## explore numeric variable that functions like a categorical variable
head(gapminder$year)
summary(gapminder$year)

## explore factor variable
class(gapminder$continent)
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
barplot(table(gapminder$continent))

## illustrative plot with ggplot2 (we'll learn this tomorrow) ----
library(ggplot2) # install.packages("ggplot2")
p <- ggplot(subset(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # initializes stuff
p <- p + scale_x_log10() # log the x axis the right way
p + geom_point() # scatterplot
p + geom_point(aes(color = continent)) # map continent to color
p + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
p + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
  geom_smooth(lwd = 1.5, se = FALSE)

## subset gapminder. Self documenting!
subset(gapminder, subset = country == "Mexico",
       select = c(country, year, lifeExp)) # ?c: combines values


## the beautiful 'then' operator %>% ----

## confusing nested way...
barplot(
  table(
    sample(
      head(gapminder$continent, 100),
      replace = FALSE,
      prob    = NULL)),
  xlab = 'continents',
  ylab = 'count')

## elegant chained way %>% 
library(magrittr) # install.packages('magrittr')
gapminder$continent %>%    # start with the data
  head(100) %>%            # before this was head(data, 100); %>% passes data
  sample(replace = FALSE,  # arguments nicely with the function name
         prob    = NULL) %>%
  table() %>%
  barplot(xlab = 'continents',
          ylab = 'count')  # no ))) or ,

```


## Review of data.frames so far

Use data.frames!!!

Important practices

  * in your analyses, do computations or make figures *in situ* -- don't create little copies and excerpts of your data. This will leave a cleaner workspace and cleaner code.
  * use %>% so that your code is easier to read and fairly self-documenting
  * give variables short informative names (`lifeExp` versus "X5")
  * refer to variables by name, not by column number
  
This will produce code that is self-documenting and more robust. Variable names often propagate to downstream outputs like figures and numerical tables and therefore good names have a positive multiplier effect throughout an analysis.

The most modern, slick way to work with data.frame is with the `dplyr` package. We will cover this tomorrow. Further reference is 

  * [Introduction to dplyr](http://stat545.com/block009_dplyr-intro.html)
  * [`dplyr` functions for a single dataset](http://stat545.com/block010_dplyr-end-single-table.html)
  * [Cheatsheet for `dplyr` joins](http://stat545.com/bit001_dplyr-cheatsheet.html), i.e. "two table" operations

